The first high-redshift changing-look quasars

### 2020-07-30 tex/MNRAS_submission/rv1/CIV_CLQs_rv1.tex
Paper RE-SUBMITTED

\iffalse
\begin{table}
  \begin{centering}
    \begin{tabular}{l r}
      \hline
      \hline
      Transition & Energy Required \\
                      &   / eV (Ryd)\\
     \hline
     \ciii $\rightarrow $ \civ  & 47.89     (3.52) \\
      \civ   $\rightarrow $ \cv &  64.49   (4.74) \\
      \cv  $ \rightarrow $ \cvi & 392.09   (28.8) \\
      \niii $\rightarrow $ \niv  & 47.44  (3.49) \\
	\niv $\rightarrow $ \nv &  77.47  (5.69)  \\
      \nv $\rightarrow $ \nvi  &  97.89   (7.19) \\
            \hline
      \hline
    \end{tabular}
    \caption{Ionisation Energies for Carbon and Nitrogen.
	These are the energies that are needed to go 
	from the species on the left and create the ion on the right. }
  \end{centering}
\end{table}
\fi

@@

 Most of the strong lines in the rest-frame near-UV spectrum of quasars are
collisionally excited, including \ciii, C\,{\sc iv} and N\,{\sc
v}.  The \heii$\lambda$1640 recombination line is among the exceptions.
}
Details of the atomic transitions and physical mechanisms that produce
strong rest-frame UV/optical lines in quasars are given in
Table~\ref{tab:atomic_lines}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%%    SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2  
%%    SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2  
%%    SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2  
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CLQ Selection and Line Measurements}
{\bf A statistical search using optical light curves to initiate spectroscopic follow-up
  observations is proving an efficient CLQ finder, and, by design, 
our data sources and overall methodology, are in common to the recent
\citet{Ross2018}, \citet{Stern2018} and \citet{Graham2020} papers.
}

@@

{\bf As described in \citet{Calderone2017}, the uncertainties in the
measured quantities are determined by two methods: a Fisher matrix
method and a Monte Carlo resampling method.  One might expect
correlations among model parameters, e.g. emission-line luminosities
and widths, potentially causing the Fisher matrix method to
underestimate the errors.  However, upon investigation, the parameters
of the involved model components (a power-law for the continuum and a
single Lorentzian profile for the emission lines) show negligible
correlations. Therefore, we use the Fisher matrix method to estimate
the uncertainties on the model parameters.  Section 2.8.3 and Appendix
B4 of \citet{Calderone2017} gives further details and compares the
Fisher matrix and Monte Carlo resampling uncertainties and
distributions.}

@@

\subsection{Virial Black Hole Masses}\label{sec:BH_masses} 
The FWHMs of quasar broad lines are related to the mass of the SMBH
powering the quasar phenomenon, and that mass is assumed to be
constant on any human timescale.  Hence it is instructive to check
whether the virial products, which are the basic quantity used to
calculate single-epoch black hole mass estimates, show any
variation. The continuum luminosity provides a proxy for the BLR
radius, and the broad-line width (FWHM) provides a proxy for the
virial velocity. This approach assumes that the broad-line region
(BLR) is virialized \citep[e.g.][]{Shen2008b, Shen2011, Calderone2017,
Mejia-Restrepo2018} and the virial mass estimate is then:
{\bf
\begin{equation}
%  \log(M) = \log \left[c\left(  \frac{\nu L_\nu}{10^{42} {\rm erg\ s^{-1}}} \right)^{\gamma}  \left( \frac{\rm FWHM}{10^3 {\rm km\ s^{-1}}} \right)^2 \right] + {\rm constant}.
  \log \left( \frac{M_{\rm BH}}{M_{\odot}} \right)  = a + \log  \left[ \; \left(  \frac{L_{1450}} {10^{44} \; {\rm erg \; s^{-1}}}\right)^{\gamma} \, \right] + 2 \log \left( \frac{\rm {FWHM}}{{\rm km\; s^{-1}}} \right)
  \label{eqn:virial_masses}
\end{equation}
where $L_{1450}$ is the monochromatic continuum luminosity at 1450\AA\,
and FWHM is the full width at half-maximum of the broad \civ\ emission line.
The coefficient $\gamma$ and offset $a$ are generally deteremined empirically.

@@

{\bf The values in Table~\ref{tab:QSFit-results} are
$\approx$0.3 dex higher than the estimates in
Table~\ref{tab:Eddington_ratios}, which is to be expected given the
difference of calibrations factors. In fact, the virial products are
all remarkably constant and compatible with a single black hole mass
estimate, within the standard 0.5 dex uncertainty, across the 3 epochs
for all the quasars.  This fact gives us confidence in the line
fitting methods.}

@@

{\bf With the black hole masses in hand, we can calculate the
Eddington luminosities of the three quasars. To be consistent with the 
bolometric luminosity, $L_{\rm bol}$, reported in \citet{Shen2011} and
\citet{Kozlowski2017} use the bolometric corrections of BC$_{1350}$ =
3.81 \citep[from the composite spectral energy distribution
in][]{Richards2006b} and to be consistent with those measurements we
adopt the same correction.  We report the the bolometric luminosities
and the Eddington ratios, $\eta=\log(L_{}/L_{\rm Edd})$, in
Table~\ref{tab:Eddington_ratios}.  }

@@

Using the same dataset as \citet{MacLeod2019}, \citet{Homan2020}
investigate the responsiveness of the \mgii\ broad emission line
doublet in AGN on timescales of several years.  Again focussing on
quasars that show large changes in their optical light-curves,
\citet{Homan2020} find that \mgii\ clearly does respond to the
continuum.  However, a key finding from \citet{Homan2020} is that the
degree of responsivity varies strikingly from one object to another.
There are cases of \mgii\ changing by as much as the continuum, more
than the continuum, or very little at all.  In the majority (72\%) of
this highly variable sample, the behaviour of \mgii\ corresponds with
that of H$\beta$.  However, there are also examples of \mgii\ showing
variations while but H$\beta$ does not, and vice versa.

@@

\subsection{Continuum and Line Changes: Comparisons to Theoretical Expectations}\label{sec:theory}
The \civ\ line is one of the strongest collisionally excited lines in
quasar spectra \citep[e.g.,][]{HamannFerland1999}.  {\bf Early
measurements from e.g., \citet{PoggePeterson1992} and
\citet{Peterson1997book}, using the well-studied Seyfert galaxy NGC
5548, suggested the location of \civ\ emission was produced by the
innermost disk, with the line emission responding to the underlying
continuum emission in only a matter of $\sim$8 days.  as indicated by
reverberation mapping time-delay measurements.  However, more recent
measurements from e.g. \citet[][and references therein]{Grier2019},
suggest that the lag between the \civ\ line and the continuum are of
order $>100$ observed-frame days.  which would place the \civ emitting
region far from the inner accretion disk.  Importantly, many quasars
in the \citet{Grier2019} sample are of comparable redshift and
luminosity to SDSS J1205+3422, J1638+2827, and J2228+2201.}

In standard \citet{SS73} thin disk models, large changes in the
continuum flux are not permitted over short timescales due to the
relatively long viscous time associated with such disks. Given the
observed short timescale continuum variations, we must consider more
complex models. We note that two of our sources (J1205+3422 and
J2228+2201) have \civ\ EW and continuum luminosity changes which fall
comfortably along a line, implying their variation is consistent with
an intrinsic Baldwin effect (see Figure~\ref{fig:CIV_Baldwin}).
%%
{\bf The light-crossing time of the region that produces the \civ\ line is substantial,
which translates into a delay between the variations of the continuum and
the \civ\ line, and a smoothing of the light curve of the latter.
%%
Thus, if the excitation of the \civ\ line is driven by the changes in the
continuum, the excited gas cannot reach an equilibrium
excitation state as quickly as the continuum changes.}
This makes the slope of the intrinsic Baldwin effect greater than that
of the overall ensemble slope, which is derived from many single-epoch
observations, the majority of which are assumed to be in equilibrium.

The presence of an intrinsic Baldwin effect implies those sources may
comfortably fit into the sample of \civ\ variable quasars explored by
\citet{Dyer2019}. Similar to those authors, we consider slim accretion
disk models \citep[e.g.,][]{Abramowicz1988}, which may explain the
observed variability. In particular, the summary of disk variation
timescales presented in \citet{Stern2018} shows that timescales are
shorter for taller disks, permitting changes similar to those observed
if they are caused by heating or cooling fronts propagating on the
disk sound crossing time. \citet{Dyer2019} also consider inhomogeneous
disk models \citep[e.g.,][]{DexterAgol2011} where flux variations are
driven by azimuthal inhomogeneities in the temperature of the
disk. Such inhomogeneities could arise from disk instabilities
\citep[e.g.,][]{LightmanEardley1974} or interactions of embedded
objects \citep[e.g.,][]{McKernan2014, McKernan2018}. If an
inhomogeneous disk is responsible for the continuum variations, which
then drive \civ\ variations, it implies that our objects are simply
the extreme outliers produced by a process which occasionally, but
rarely, produces very large hot or cool spots. The frequency of such
occurrences can be used to constrain the slope and normalization of
the power law distribution of spot size.

@@

The foregoing discussion directly applies to the two objects which
maintain an intrinsic Baldwin effect relationship between \civ\ EW and
continuum luminosity. However, J1638+2827 clearly does not maintain
such a relationship --- indeed, the continuum and \civ\ EW appear
somewhat correlated, rather than anti-correlated, in this source. This
implies that the illumination of the \civ\ emitting region by a
variable ionising continuum and the corresponding change in the
photoionisation state alone cannot explain the collapse and recovery
of the line.

@@

\citet{NodaDone2018} note that the soft X-ray excess -- the excess of
X-rays below 2 keV with respect to the extrapolation of the hard X-ray
spectral continuum model, and a common feature among Type~1 AGN --
produces most of the ionizing photons, so its dramatic drop can lead
to the disappearance of the broad-line region, driving the
changing-look phenomena. \citet{NodaDone2018} go on to make a
connection between state changes in Galactic binaries and CLQs.  In
Galactic binaries spectral hardening corresponds to the soft-to-hard
state transition in black hole binaries, which occurs at $L / L_{\rm
Edd} \sim 0.02$ (i.e., $\eta_{\rm Edd} \sim -1.7$).  During this state
transition, the inner disc evaporates into an advection-dominated
accretion flow, while the overall drop in luminosity appears
consistent with the hydrogen ionisation disc instability.
\citet{NodaDone2018} predict that changing-look AGNs are similarly
associated with state transitions at $L / L_{\rm Edd}$ of a few
percent.

@@

Apart from J1205+3422 on MJD 58538, we see all the \civ\ CLQ epochs
are above $L / L_{\rm Edd} \sim$5\%. As we have noted throughout, the
spectrum for J1205+3422 on MJD 58538 has low SNR, so while this object
could well have entered a `low-state', this is difficult to
conclusively confirm.  Nevertheless, and taking the J1205+3422
spectrum at face-value, we note that all the \civ\ CLQs are above
$\sim$1\% in Eddington luminosity and thus the suggested ``Low/Hard
State'' boundary.

While we note there is interest in plotting these accretion rates, and
motivation from \citet{NodaDone2018} and \citet{Ruan2019a}, we very
much caution on over-interpretation at this juncture. The point of
this paper was to report on three very interesting CLQs.  Our object
definitions are based on empirical, observed properties.  The
potential for physical connections of accretion physics across the
large-range of mass-scales is tantalising, but is left to future
investigations.

@@

There are now more observed examples of dramatic, dynamic changes in
supermassive black hole systems e.g. CLQs/CSQs than Galactic X-ray
Binaries.  While the timescales are expected to scale with black hole
mass, whether the full physical system at large does, including the
associated atomic physics, remains an open question.

@@

%NPR dedicates this work, and indeed his previous research to his father. Dad had a deep love for science, mathematics and questioning the conventional wisdom. His support allowed me to become an astrophysicist and to have the adventures and discoveries we did. He is sorely missed.

>>>>

@@

Most of the strong lines in the rest-frame near-UV spectrum of quasars
are collisionally excited, including \ciii, C\,{\sc iv} and N\,{\sc
v}.  The \heii$\lambda$1640 recombination line is among the
exceptions.} Details of the atomic transitions and physical mechanisms
that produce strong rest-frame UV/optical lines in quasars are given
in Table~\ref{tab:atomic_lines}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%%    SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2  
%%    SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2  
%%    SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2   SECTION 2  
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{CLQ Selection and Line Measurements}
{\bf A statistical search using optical light curves to initiate
spectroscopic follow-up observations is proving an efficient CLQ
finder, and, by design, our data sources and overall methodology, are
in common to the recent \citet{Ross2018}, \citet{Stern2018} and
\citet{Graham2020} papers.}

@@

{\bf As described in \citet{Calderone2017}, the uncertainties in the
measured quantities are determined by two methods: a Fisher matrix
method and a Monte Carlo resampling method.  One might expect
correlations among model parameters, e.g. emission-line luminosities
and widths, potentially causing the Fisher matrix method to
underestimate the errors.  However, upon investigation, the parameters
of the involved model components (a power-law for the continuum and a
single Lorentzian profile for the emission lines) show negligible
correlations. Therefore, we use the Fisher matrix method to estimate
the uncertainties on the model parameters.}

@@

\subsection{Virial Black Hole Masses}\label{sec:BH_masses} 
The FWHMs of quasar broad lines are related to the mass of the SMBH
powering the quasar phenomenon, and that mass is assumed to be
constant on any human timescale.  Hence it is instructive to check
whether the virial products, which are the basic quantity used to
calculate single-epoch black hole mass estimates, show any
variation. The continuum luminosity provides a proxy for the BLR
radius, and the broad-line width (FWHM) provides a proxy for the
virial velocity. This approach assumes that the broad-line region
(BLR) is virialized \citep[e.g.][]{Shen2008b, Shen2011, Calderone2017,
Mejia-Restrepo2018} and the virial mass estimate is then:
{\bf
\begin{equation}
  \log \left( \frac{M_{\rm BH}}{M_{\odot}} \right)  = a + \log  \left[ \; \left(  \frac{L_{1450}} {10^{44} \; {\rm erg \; s^{-1}}}\right)^{\gamma} \, \right] + 2 \log \left( \frac{\rm {FWHM}}{{\rm km\; s^{-1}}} \right)
  \label{eqn:virial_masses}
\end{equation}
where $L_{1450}$ is the monochromatic continuum luminosity at
1450\AA\, and FWHM is the full width at half-maximum of the broad
\civ\ emission line.  The coefficient $\gamma$ and offset $a$ are
generally deteremined empirically.

@@

{\bf The values in Table~\ref{tab:QSFit-results} are $\approx$0.3 dex
higher than the estimates in Table~\ref{tab:Eddington_ratios}, which
is to be expected given the difference of calibrations factors. In
fact, the virial products are all remarkably constant and compatible
with a single black hole mass estimate, within the standard 0.5 dex
uncertainty, across the 3 epochs for all the quasars.}

@@

{\bf With the black hole masses in hand, we can calculate the
Eddington luminosities of the three quasars. To be consistent with the
bolometric luminosity, $L_{\rm bol}$, reported in \citet{Shen2011} and
\citet{Kozlowski2017} use the bolometric corrections of BC$_{1350}$ =
3.81 \citep[from the composite spectral energy distribution
in][]{Richards2006b} and to be consistent with those measurements we
adopt the same correction.  We report the the bolometric luminosities
and the Eddington ratios, $\eta=\log(L_{}/L_{\rm Edd})$, in
Table~\ref{tab:Eddington_ratios}.}

@@

Using the same dataset as \citet{MacLeod2019}, \citet{Homan2020}
investigate the responsiveness of the \mgii broad emission line
doublet in AGN on timescales of several years.  Again focussing on
quasars that show large changes in their optical light-curves,
\citet{Homan2020} find that \mgii clearly does respond to the
continuum.  However, a key finding from \citet{Homan2020} is that the
degree of responsivity varies strikingly from one object to another.
There are cases of \mgii\ changing by as much as the continuum, more
than the continuum, or very little at all.  In the majority (72\%) of
this highly variable sample, the behaviour of \mgii corresponds with
that of H$\beta$.  However, there are also examples of \mgii showing
variations while H$\beta$ does not, and vice versa.

@@

\subsection{Continuum and Line Changes: Comparisons to Theoretical Expectations}\label{sec:theory}
The \civ\ line is one of the strongest collisionally excited lines in
quasar spectra \citep[e.g.,][]{HamannFerland1999}.
%%
{\bf Early measurements from e.g., \citet{PoggePeterson1992} and
\citet{Peterson1997book}, using the well-studied Seyfert galaxy NGC
5548, suggested the location of \civ\ emission was produced by the
innermost disk, with the line emission responding to the underlying
continuum emission in only a matter of $\sim$8 days, as indicated by
reverberation mapping time-delay measurements. NGC 5548 has an SMBH
mass of $\sim$70$\times 10^{6} M_{\odot}$ \citep[][]{Horne2020} and
the bolometric luminosity of NGC 5548 is between $\sim$0.2-10\% of the
Eddington luminosity \citep{Ebrero2016, Fausnaugh2016}, giving a
$\log(L_{\rm bol, NGC5448}$/ erg s$^{-1}$) = $\sim$43.2 - 44.9.
Thus, the CIV CLQs have a central SMBH mass of $\sim$7-90$\times$ that
of NGC 5548, and a bolometric luminosity $\sim$3-1400$\times$
greater. Quasars in the recent \citet{Grier2019} study are of
comparable redshift and luminosity to J1205+3422, J1638+2827, and
J2228+2201. \citet{Grier2019} measure the lag between the \civ\ line
and the continuum to be of order $>100$ observed-frame days.}

In standard \citet{SS73} thin disk models, large changes in the
continuum flux are not permitted over short timescales due to the
relatively long viscous time associated with such disks. Given the
observed short timescale continuum variations, we must consider more
complex models. We note that one of our sources (J1205+3422) has \civ\
EW and continuum luminosity changes which fall comfortably along a
line, implying their variation is consistent with an intrinsic Baldwin
effect (see Figure~\ref{fig:CIV_Baldwin}).
{\bf The light-crossing time of the region that produces the \civ\
line is substantial, which translates into a delay between the
variations of the continuum and the \civ\ line, and a smoothing of the
light curve of the latter.  Thus, if the excitation of the \civ\ line
is driven by the changes in the continuum, the excited gas cannot
reach an equilibrium excitation state as quickly as the continuum
changes.}
This makes the slope of the intrinsic Baldwin effect greater than that
of the overall ensemble slope, which is derived from many single-epoch
observations, the majority of which are assumed to be in equilibrium.

The presence of an intrinsic Baldwin effect implies those sources may
comfortably fit into the sample of \civ\ variable quasars explored by
\citet{Dyer2019}. Similar to those authors, we consider slim accretion
disk models \citep[e.g.,][]{Abramowicz1988}, which may explain the
observed variability. In particular, the summary of disk variation
timescales presented in \citet{Stern2018} shows that timescales are
shorter for taller disks, permitting changes similar to those observed
if they are caused by heating or cooling fronts propagating on the
disk sound crossing time. \citet{Dyer2019} also consider inhomogeneous
disk models \citep[e.g.,][]{DexterAgol2011} where flux variations are
driven by azimuthal inhomogeneities in the temperature of the
disk. Such inhomogeneities could arise from disk instabilities
\citep[e.g.,][]{LightmanEardley1974} or interactions of embedded
objects \citep[e.g.,][]{McKernan2014, McKernan2018}. If an
inhomogeneous disk is responsible for the continuum variations, which
then drive \civ\ variations, it implies that J1205+3422 is simply an
extreme outlier produced by a process which occasionally, but rarely,
produces very large hot or cool spots. The frequency of such
occurrences can be used to constrain the slope and normalization of
the power law distribution of spot size.

@@

The foregoing discussion directly applies to J1205+3422, which
maintains a clear intrinsic Baldwin effect relationship between \civ\
EW and continuum luminosity (J2228+2201 may also be consistent with
this picture, given the measurement uncertainties).  However,
J1638+2827 clearly does not maintain such a relationship --- indeed,
the continuum and \civ\ EW appear somewhat correlated, rather than
anti-correlated, in this source. This implies that the illumination of
the \civ\ emitting region by a variable ionising continuum and the
corresponding change in the photoionisation state alone cannot explain
the collapse and recovery of the line.

@@

\citet{NodaDone2018} note that the soft X-ray excess -- the excess of
X-rays below 2 keV with respect to the extrapolation of the hard X-ray
spectral continuum model, and a common feature among Type~1 AGN --
produces most of the ionizing photons, so its dramatic drop can lead
to the disappearance of the broad-line region, driving the
changing-look phenomena. \citet{NodaDone2018} go on to make a
connection between state changes in Galactic binaries and CLQs.  In
Galactic binaries, spectral hardening corresponds to the soft-to-hard
state transition in black hole binaries, which occurs at $L / L_{\rm
Edd} \sim 0.02$ (i.e., $\eta_{\rm Edd} \sim -1.7$).  During this state
transition, the inner disc evaporates into an advection-dominated
accretion flow, while the overall drop in luminosity appears
consistent with the hydrogen ionisation disc instability.
\citet{NodaDone2018} predict that changing-look AGNs are similarly
associated with state transitions at $L / L_{\rm Edd}$ of a few
percent.

@@

Apart from J1205+3422 on MJD 58538 (the observation obtained in poor
conditions) we see all the \civ\ CLQ epochs are above $L / L_{\rm Edd}
\sim$5\%. The spectrum for J1205+3422 on MJD 58538 has low SNR, so
while this object could well have entered a `low-state', this is
difficult to conclusively confirm.  Nevertheless, and taking the
J1205+3422 spectrum at face-value, we note that all the \civ\ CLQs are
above $\sim$1\% in Eddington luminosity and thus above the suggested
``Low/Hard State'' boundary.

While we note there is interest in plotting these accretion rates, and
motivation from \citet{NodaDone2018} and \citet{Ruan2019a}, we caution
against over-interpretation at this juncture. The point of this paper
was to report on three very interesting CLQs.  Our object definitions
are based on empirical, observed properties.  The potential for
physical connections of accretion physics across the large-range of
mass-scales is tantalising, but is left to future investigations.

@@

There are now more observed examples of dramatic, dynamic changes in
supermassive black hole systems, e.g. CLQs/CSQs, than there are
Galactic X-ray Binaries.  While the timescales are expected to scale
with black hole mass, whether the full physical system at large does,
including the associated atomic physics, remains an open question.

@@

NPR dedicates this work, and indeed his previous research, to his
father. Dad had a deep love for science, mathematics and questioning
the conventional wisdom. His support allowed me to become an
astrophysicist and to have the adventures and discoveries we did. He
is sorely missed.
